# CYCLISTIC BIKE SHARE ANALYSIS #

#### This analysis is for a case study from the Google Data Analytics Certificate ####

#### The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: ####

#### In what ways do members and casual riders use Divvy bikes differently? ####

### STEP-1: Installation of required packages & change of directory for ease. ###
```{r Packages Installation & directory setting, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse) #helps wrangle data
library(lubridate) #helps weangle data attributes
library(ggplot2) #helps to visualize data
getwd() #displays your working directory
setwd("C:/Users/TEMP/Downloads/GDA/Cyclistic Bike Share") #sets your working directory to simplify calls to data
```
### STEP-2: Data import into R. ###
```{r Data Import, message=FALSE, warning=FALSE, paged.print=FALSE}
#read csv() imports data from .csv files.
Jan_2021 <- read_csv("202101-divvy-tripdata.csv")
Feb_2021 <- read_csv("202102-divvy-tripdata.csv")
Mar_2021 <- read_csv("202103-divvy-tripdata.csv")
Apr_2020 <- read_csv("202004-divvy-tripdata.csv")
May_2020 <- read_csv("202005-divvy-tripdata.csv")
Jun_2020 <- read_csv("202006-divvy-tripdata.csv")
Jul_2020 <- read_csv("202007-divvy-tripdata.csv")
Aug_2020 <- read_csv("202008-divvy-tripdata.csv")
Sep_2020 <- read_csv("202009-divvy-tripdata.csv")
Oct_2020 <- read_csv("202010-divvy-tripdata.csv")
Nov_2020 <- read_csv("202011-divvy-tripdata.csv")
Dec_2020 <- read_csv("202012-divvy-tripdata.csv")
```
### STEP 3: Comparing column names for each of the files. ###
```{r Column Names Check, message=FALSE, warning=FALSE, paged.print=FALSE}
#colnames() displays the names of all the columns in the dataset.
colnames(Jan_2021)
colnames(Feb_2021)
colnames(Mar_2021)
colnames(Apr_2020)
colnames(May_2020)
colnames(Jun_2020)
colnames(Jul_2020)
colnames(Aug_2020)
colnames(Sep_2020)
colnames(Oct_2020)
colnames(Nov_2020)
colnames(Dec_2020)
```
### STEP-4: Inspecting the dataframes and look for incongruencies ###
```{r Data Frames Inspection, message=FALSE, warning=FALSE, paged.print=FALSE}
#str() gives the data type info of data in the dataset.
str(Jan_2021)
str(Feb_2021)
str(Mar_2021)
str(Apr_2020)
str(May_2020)
str(Jun_2020)
str(Jul_2020)
str(Aug_2020)
str(Sep_2020)
str(Oct_2020)
str(Nov_2020)
str(Dec_2020)
```
### STEP-5: Making consistent Data types for columns in each of the datasets ###
####  Because they need to match perfectly before we can use a command to join them into one file ####
```{r}
#mutate() is used to change the data type of a column from one to another.
Jan_2021 <-  mutate(Jan_2021, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Feb_2021 <-  mutate(Feb_2021, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Mar_2021 <-  mutate(Mar_2021, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Apr_2020 <-  mutate(Apr_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id))  
May_2020 <-  mutate(May_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id))  
Jun_2020 <-  mutate(Jun_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Jul_2020 <-  mutate(Jul_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Aug_2020 <-  mutate(Aug_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Sep_2020 <-  mutate(Sep_2020, start_station_id = as.double(start_station_id)
                   ,end_station_id = as.double(end_station_id)) 
Oct_2020 <-  mutate(Oct_2020, start_station_id = as.double(start_station_id)
                    ,end_station_id = as.double(end_station_id))  
Nov_2020 <-  mutate(Nov_2020, start_station_id = as.double(start_station_id)
                    ,end_station_id = as.double(end_station_id))  
Dec_2020 <-  mutate(Dec_2020, start_station_id = as.double(start_station_id)
                    ,end_station_id = as.double(end_station_id))
```
### STEP-6: Combining all the datasets into one Data Frame ###
```{r}
#bind_rows combines rows from various datasets into one data frame.
all_trips <- bind_rows(Jan_2021, Feb_2021, Mar_2021, Apr_2020, May_2020, Jun_2020, Jul_2020, Aug_2020, Sep_2020, Oct_2020, Nov_2020, Dec_2020)
```
### STEP-7: Removal of Unnecessary Columns from the Data Frame ###
```{r}
#select(-argument) remvoes selected columns from the dataset.
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```
### STEP-8: Insight gathering from the Data Frame ###
```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```
### STEP-9: Seperating Date, Month, Year & Day for better data aggregation ###
```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
### STEP-10: Calculating Trip DUration for each ride ###
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```
### STEP-11: Remvoing Irrelevant/Incomplete Data ###
#### Made a new version of the file incase we need to comeback to the data ####
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
### STEP-12: Descriptive analysis on ride_length (all figures in seconds) ###
```{r}
summary(all_trips_v2$ride_length)
```
### STEP-13: Comparing Member and casual users (all figures in seconds) ###
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
### STEP-14: Average ride time by each day for members vs casual users (all figures in seconds) ###
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
### STEP-15: Analyzing riders data by type and weekday (all figures in seconds) ###
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```
### STEP-16: Visualizing the number of rides in each weekday by member type ###
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides by Member type during eachday of the week")
```
### STEP-17: Visualization for average duration of rides in each weekday by member type ###
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average duration of rides in each weekday by member type")
```
### STEP-18: Visualization for number of rides in each ride type by member types ###
```{r}
all_trips_v2 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual)  %>% 
  ggplot(aes(x = rideable_type, y=number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides in each ride type by member types")
```
### STEP-19: Saving a new file for further Analysis ###
```{r}
write.csv(all_trips_v2, file = 'C:/Users/TEMP/Downloads/GDA/Cyclistic Bike Share/avg_ride_length_1.csv')
```